---
title: "Homework 04: Tidy data and joins"
author: Emre Erhan
date: October 9, 2018
output:
  github_document:
    toc: yes 
---

```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(dplyr))
```

## Make a tibble with one row per year and columns for life expectancy for two or more countries. 

```{r}
tidy_lifeExp_data = gapminder %>%
  filter(country %in% c('United States', 'Canada', 'Turkey')) %>%
  select(year, lifeExp, country) %>%
  spread(country, lifeExp)
knitr::kable(tidy_lifeExp_data)
```

```{r}
tidy_lifeExp_data %>%
  ggplot(aes(x = Canada, y = `United States`)) +
  ggtitle("The Life Expectancy of Canada vs USA") +
  geom_point() +
  geom_smooth()
  
```

## Explore database style joining in R with my `nato` `data.frame`

```{r}
# The following csv file was manually copied line-by-line from https://en.wikipedia.org/wiki/Member_states_of_NATO
nato = read.csv('nato-countries.csv')
```

### Let's start by looking at outer joins.

The following left join includes all columnns from `nato`, but only contains values for countries belonging to NATO. As you can see from the `head` of the joined `data.frame`, there are no `dateJoined` and `enlargment` values for Afghanistan, since the country is not found in the `nato` `data.frame`.

```{r}
gapminder_nato_left_join = left_join(gapminder, nato, by = "country") # An SQL style left join
head(gapminder_nato_left_join) %>%
  knitr::kable()
```

The following code only looks at the countries that are found in `nato`.

```{r}
gapminder_nato_right_join = right_join(gapminder, nato, by = "country") # An SQL style right join
head(gapminder_nato_right_join) %>%
  knitr::kable()
```

### Let's try an `anti_join` to see if `gapminder` is missing any NATO member countries.

```{r}
anti_join(nato, gapminder, by = "country") %>%
  knitr::kable()
```

Interestingly, the above five Eastern European countries that are missing from the `gapminder` dataset.

### Let's combine our knowledge, and see how the `pop` has changed for different `enlargement`s of NATO member countries

```{r}
gapminder_nato_right_join %>%
  select(enlargement, year, lifeExp) %>%
  group_by(year, enlargement) %>%
  summarise(lifeExp = mean(lifeExp)) %>%
  arrange(enlargement) %>%
  ggplot(aes(year, lifeExp, colour = enlargement)) +
    ggtitle("Examining the change in mean life expectation of NATO countries by Enlargement") +
    geom_line()
```